---
title: "final data exploration"
author: "shawn"
date: "03/04/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Includes packages and read in the data set
```{r}
#read in pollutants data set
pollutants <- read.csv("C:/Users/Shawn/Downloads/pollutants.csv")
library(car)
library(ggplot2)
library(ggcorrplot)
library(caret)
library(glmnet)
library(MASS)
library(tidyverse)
library(rmarkdown)
library(olsrr)
```

##Data set cleaning
```{r}
#remove the x
pollutants$X = NULL
```

##Summary of the dataset
```{r}
#summary of data
summary(pollutants)
```


##Getting the Correlation Matrix Heat Map
```{r}
#calculate correlation matrix before removing the multicolinearity covariates
corr_mat = cor(pollutants)
#graph colored corr matrix
ggcorrplot(corr_mat, legend.title = "Correlation plot with Categoricals")
```

##get Correlation Matrix Heat map without categorical
```{r}
#get set a dataset with no categorical covariates
no_cat = pollutants
no_cat$edu_cat = NULL
no_cat$race_cat = NULL
no_cat$male = NULL
no_cat$smokenow = NULL
#calculate correlation matrix
corr_matrix = cor(no_cat)
#graph colored corr matrix
ggcorrplot(corr_matrix, legend.title = "Correlation plot with no Categoricals")
```

##Graph the Boxplot and Histograms
```{r}
#graph boxplotograms
boxplot(pollutants$length, xlab = "length", ylab = "Value")
boxplot(pollutants$POP_PCB1, xlab = "POP_PCB1", ylab = "Value")
boxplot(pollutants$POP_PCB2, xlab = "POP_PCB2", ylab = "Value")
boxplot(pollutants$POP_PCB3, xlab = "POP_PCB3", ylab = "Value")
boxplot(pollutants$POP_PCB4, xlab = "POP_PCB4", ylab = "Value")
boxplot(pollutants$POP_PCB5, xlab = "POP_PCB5", ylab = "Value")
boxplot(pollutants$POP_PCB6, xlab = "POP_PCB6", ylab = "Value")
boxplot(pollutants$POP_PCB7, xlab = "POP_PCB7", ylab = "Value")
boxplot(pollutants$POP_PCB8, xlab = "POP_PCB8", ylab = "Value")
boxplot(pollutants$POP_PCB9, xlab = "POP_PCB9", ylab = "Value")
boxplot(pollutants$POP_PCB10, xlab = "POP_PCB10", ylab = "Value")
boxplot(pollutants$POP_PCB11, xlab = "POP_PCB11", ylab = "Value")
boxplot(pollutants$POP_dioxin1, xlab = "POP_dioxin", ylab = "Value")
boxplot(pollutants$POP_dioxin2, xlab = "POP_dioxin2", ylab = "Value")
boxplot(pollutants$POP_dioxin3, xlab = "POP_dioxin3", ylab = "Value")
boxplot(pollutants$POP_furan1, xlab = "POP_furan1", ylab = "Value")
boxplot(pollutants$POP_furan2, xlab = "POP_furan2", ylab = "Value")
boxplot(pollutants$POP_furan3, xlab = "POP_furan3", ylab = "Value")
boxplot(pollutants$POP_furan4, xlab = "POP_furan4", ylab = "Value")
boxplot(pollutants$whitecell_count, xlab = "whitecell_count", ylab = "Value")
boxplot(pollutants$lymphocyte_pct, xlab = "lymphocyte_pct", ylab = "Value")
boxplot(pollutants$monocyte_pct, xlab = "monocyte_pct", ylab = "Value")
boxplot(pollutants$eosinophils_pct, xlab = "eosinophils_pct", ylab = "Value")
boxplot(pollutants$basophils_pct, xlab = "basophils_pct", ylab = "Value")
boxplot(pollutants$neutrophils_pct, xlab = "neutrophils_pct", ylab = "Value")
boxplot(pollutants$BMI, xlab = "BMI", ylab = "Value")
hist(pollutants$edu_cat)
hist(pollutants$race_cat)
hist(pollutants$male)
boxplot(pollutants$ageyrs, xlab = "ageyrs", ylab = "Value")
boxplot(pollutants$yrssmoke, xlab = "yrssmoke", ylab = "Value")
hist(pollutants$smokenow)
boxplot(pollutants$ln_lbxcot, xlab = "ln_lbxcot", ylab = "Value")
```


##Rename categoricals and set factors
```{r}
#turn categorical numbers into strings
pollutants$male[pollutants$male == 1] = "Male"
pollutants$male[pollutants$male == 0] = "Female"

pollutants$smokenow[pollutants$smokenow == 1] = "Smokes"
pollutants$smokenow[pollutants$smokenow == 0] = "No Smokes"

pollutants$edu_cat[pollutants$edu_cat == 1] = "12 grade no dip below"
pollutants$edu_cat[pollutants$edu_cat == 2] = "High School Grad/GED,Eqv"
pollutants$edu_cat[pollutants$edu_cat == 3] = "Some College,AA degree"
pollutants$edu_cat[pollutants$edu_cat == 4] = "College graduate"

pollutants$race_cat[pollutants$race_cat == 1] = "Other race, mutli-racial"
pollutants$race_cat[pollutants$race_cat == 2] = "Mexican American"
pollutants$race_cat[pollutants$race_cat == 3] = "Non-Hispanic Black"
pollutants$race_cat[pollutants$race_cat == 4] = "Non-Hispanic White"

#factor the categorical
pollutants$male = as.factor(pollutants$male)
pollutants$smokenow = as.factor(pollutants$smokenow)
pollutants$edu_cat = as.factor(pollutants$edu_cat)
pollutants$race_cat = as.factor(pollutants$race_cat)
```

##See the effects of categorical on the response
```{r}
#see how length is different across Currently Smokes (Smokes) and currently does not smokes (No Smokes)
ggplot(pollutants, aes(x = male, y = length, color = male)) +
  geom_boxplot()+
  geom_point(size = 2, position = position_jitter(width = 0.2)) +
  stat_summary(fun = mean, geom = "point", shape = 20, size = 6, color = "blue")+
  theme_classic() +
  facet_grid(.~smokenow)

#see how length is different across education
ggplot(pollutants, aes(x = male, y = length, color = male)) +
  geom_boxplot()+
  geom_point(size = 2, position = position_jitter(width = 0.2)) +
  stat_summary(fun = mean, geom = "point", shape = 20, size = 6, color = "blue")+
  theme_classic() +
  facet_grid(.~edu_cat)

#see how length is different across races
ggplot(pollutants, aes(x = male, y = length, color = male)) +
  geom_boxplot()+
  geom_point(size = 2, position = position_jitter(width = 0.2)) +
  stat_summary(fun = mean, geom = "point", shape = 20, size = 6, color = "blue")+
  theme_classic() +
  facet_grid(.~race_cat)
```

##Remove VIF
```{r}
model = lm(length ~ .,data = pollutants)
#show the VIF
vif(model)


#remove covariates with VIF > 10 below:

#remove highest vif ~ 15047 
pollutants$eosinophils_pct = NULL
#fit new model
model = lm(length ~ .,data = pollutants)
#show the VIF
vif(model)

#remove highest vif eosinophils_pct ~ 15047 then POP_PCB5 ~ 59.718357
pollutants$POP_PCB5 = NULL
#fit new model
model = lm(length ~ .,data = pollutants)
#show the VIF
vif(model)

#remove highest vif eosinophils_pct ~ 15047 then POP_PCB5 ~ 59.718357 then POP_PCB1 ~ 32.255513
pollutants$POP_PCB1 = NULL
#fit new model
model = lm(length ~ .,data = pollutants)
#show the VIF
vif(model)

#remove highest vif eosinophils_pct ~ 15047 then POP_PCB5 ~ 59.718357 then POP_PCB1 ~ 32.255513 then POP_PCB2 ~ 22.564616
pollutants$POP_PCB2 = NULL
#fit new model
model = lm(length ~ .,data = pollutants)
#show the VIF
vif(model)

#remove highest vif eosinophils_pct ~ 15047 then POP_PCB5 ~ 59.718357 then POP_PCB1 ~ 32.255513 then POP_PCB2 ~ 22.564616
# then POP_PCB4 ~ 14.260201
pollutants$POP_PCB4 = NULL
#fit new model
model = lm(length ~ .,data = pollutants)
#show the VIF
vif(model)
```

##Scatterplot of the rest of the covariate
```{r}
#scatterplot after removing the high VIF covariates
pdf("scatter.pdf", height=1000, width=1000)
pairs(pollutants, cex.labels = 40, gap = 0.5)
dev.off()
knitr::include_graphics("scatter.pdf")
#You maybe have to open this scatter.pdf in the files section of the RStudio and open the scatter.pdf with browser like firefox to be 
# able to see the entire matrix because the amount of covariates there are. ~ 31MB on my PC.
```

##See how the model fits
```{r}
#find the model fit for homoscedasticity before remove multicolinearity
par(mfrow=c(2,2))
plot(model)
par(mfrow=c(1,1))
#actual vs predicted plot
plot(predict(model), pollutants$length, xlab = "Predicted", ylab = "Actual")
abline(a=0,b=1)
outlierTest(model)
#studentized Residual Plot
ols_plot_resid_stud(model)$data
#high leverage plot
ols_plot_resid_lev(model)$data
#find the cook distance plot bar plot
ols_plot_cooksd_bar(model)$data
#DFBETA
ols_plot_dfbetas(model)
#DEFITS
ols_plot_dffits(model)$data
```

#find the best model with no interaction AIC
```{r}
stepwise_AIC = step(model, direction = "both", trace = FALSE)
summary(stepwise_AIC)
#See how the model assumptions fit
par(mfrow=c(2,2))
plot(stepwise_AIC)
par(mfrow=c(1,1))
#actual vs predicted plot
plot(predict(stepwise_AIC), pollutants$length, xlab = "Predicted", ylab = "Actual")
abline(a=0,b=1)
outlierTest(stepwise_AIC)
#studentized Residual Plot
ols_plot_resid_stud(stepwise_AIC)$data
#high leverage plot
ols_plot_resid_lev(stepwise_AIC)$data
#find the cook distance plot bar plot
ols_plot_cooksd_bar(stepwise_AIC)$data
#DFBETA
ols_plot_dfbetas(stepwise_AIC)
#DEFITS
ols_plot_dffits(stepwise_AIC)$data
```

#find the best model with no interaction BIC
```{r}
stepwise_BIC = step(model, direction = "both", trace = FALSE, k=log(nrow(pollutants)))
summary(stepwise_BIC)
#See how the model assumptions fit
par(mfrow=c(2,2))
plot(stepwise_BIC)
par(mfrow=c(1,1))
#actual vs predicted plot
plot(predict(stepwise_BIC), pollutants$length, xlab = "Predicted", ylab = "Actual")
abline(a=0,b=1)
outlierTest(stepwise_BIC)
#studentized Residual Plot
ols_plot_resid_stud(stepwise_BIC)$data
#high leverage plot
ols_plot_resid_lev(stepwise_BIC)$data
#find the cook distance plot bar plot
ols_plot_cooksd_bar(stepwise_BIC)$data
#DFBETA
ols_plot_dfbetas(stepwise_BIC)
#DEFITS
ols_plot_dffits(stepwise_BIC)$data
```

##Find the best model including interactions using backward stepwise elimination AIC
```{r}
aic_matrix = as.data.frame(model.matrix(stepwise_AIC))
aic_matrix = aic_matrix[, -1]
length = pollutants$length
interact_aic = lm(length ~ .^2,data = aic_matrix)
AIC_result = step(interact_aic, direction = "backward", trace = FALSE, k=2)
summary(AIC_result)
#See how the model assumptions fit
par(mfrow=c(2,2))
plot(AIC_result)
par(mfrow=c(1,1))
#actual vs predicted plot
plot(predict(AIC_result), pollutants$length, xlab = "Predicted", ylab = "Actual")
abline(a=0,b=1)
outlierTest(AIC_result)
#studentized Residual Plot
ols_plot_resid_stud(AIC_result)$data
#high leverage plot
ols_plot_resid_lev(AIC_result)$data
#find the cook distance plot bar plot
ols_plot_cooksd_bar(AIC_result)$data
#DFBETA
ols_plot_dfbetas(AIC_result)
#DEFITS
ols_plot_dffits(AIC_result)$data
```

##Find the best model including interactions using backward stepwise elimination BIC
```{r}
bic_matrix = as.data.frame(model.matrix(stepwise_BIC))
bic_matrix = bic_matrix[, -1]
length = pollutants$length
interact_bic = lm(length ~ .^2,data = bic_matrix)
BIC_result = step(interact_bic, direction = "backward", trace = FALSE, k=log(nrow(pollutants)))
summary(BIC_result)
#See how the model assumptions fit
par(mfrow=c(2,2))
plot(BIC_result)
par(mfrow=c(1,1))
#actual vs predicted plot
plot(predict(BIC_result), pollutants$length, xlab = "Predicted", ylab = "Actual")
abline(a=0,b=1)
outlierTest(BIC_result)
#studentized Residual Plot
ols_plot_resid_stud(BIC_result)$data
#high leverage plot
ols_plot_resid_lev(BIC_result)$data
#find the cook distance plot bar plot
ols_plot_cooksd_bar(BIC_result)$data
#DFBETA
ols_plot_dfbetas(BIC_result)
#DEFITS
ols_plot_dffits(BIC_result)$data
```
